---
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    template: null
#    citation_package: natbib
biblio-style: naturemag-doi
bibliography: ppmiscidata.bib
classoption: table
header-includes:
  - \usepackage{multirow}
  - \usepackage{rotating}
  - \usepackage{authblk}
  - \renewcommand\Affilfont{\small}
  - \usepackage{booktabs}
  - \usepackage{lscape}
linenumbers: true
urlcolor: blue
# nocite: '@*'
link-citations: true
---



\begin{centering}

$ $

\vspace{0.1 cm}

\LARGE

{\bf Magnetic Resonance Imaging Data Phenotypes for the Parkinson's Progression Markers Initiative: Supplemental information}

\vspace{1.0 cm}

\normalsize

Brian B Avants$^{1,2}$,
Leon Fonville$^{1}$,
Olivia Hampton$^{1}$,
Alexandra Reardon$^{1}$,
Andrew Stenger$^{1}$,
Xue Wang$^{1}$,
Nicholas J Tustison$^{2}$,
James R Stone$^{2}$,
Philip A Cook$^3$,
Barbara Marebwa$^4$,
Lana M Chahine$^5$,
Kathleen L Poston$^6$,
Kenneth Marek$^7$,
Lino Becerra$^{1}$,
Roger Gunn$^{1}$
for the Alzheimer's Disease Neuroimaging Initiative*


\vspace{0.25 cm}

\small

$^1$Invicro, Needham, MA, USA

$^2$Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA, USA

$^3$Department of Radiology, Perelman School of Medicine at the University of Pennsylvania, Philadelphia, PA, USA

$^4$Parkinson's Research, The Michael J. Fox Foundation,  New York City, New York, USA

$^5$Department of Neurology, University of Pittsburgh, Pittsburgh, PA, USA

$^6$Department of Neurology, Stanford University, Palo Alto, CA, USA

$^7$Institute for Neurodegenerative Disorders, New Haven, CT, USA


\end{centering}

\vspace{2.2 cm}

\small

Corresponding author: \
Brian B. Avants \
avants@grasp.cis.upenn.edu

\noindent\rule{4cm}{0.4pt}

\footnotesize

*Data used in preparation of this article were obtained from the Alzheimer’s Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at: http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf

\newpage

\normalsize


\LaTeX{
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
}


```{r samplesizes,echo=FALSE,eval=TRUE,warning=FALSE}
redomerge=TRUE
buildsci=buildsci2=TRUE
```


```{r ppmimerge, echo=FALSE,eval=redomerge,cache=FALSE,warning=FALSE}
library(ztable)
library(rsq)
options(ztable.zebra.color="platinum")
options(ztable.colnames.bold=TRUE)
options(ztable.include.rownames=FALSE)
library(moonBook)
library(plyr)
library(subtyper)
library(forcats)
library(lmerTest)
library(lme4)
library(ggplot2)
library(ggpubr)
library(gridExtra)
library(mclust)
library(MASS)
library(dCUR)
library(mlr3learners)
library(mlr3verse)
library(effects)
library(ANTsR)
library(glossary)
glossary_reset()
glosspath='/Users/stnava/Documents/writing/ppmi_mri_scidata/data/ppmi_glossary.yml'
glossary_path(glosspath)
glossary_add(
  term = "Parkinson's disease",
  def = "A progressive neurodegenerative disorder characterized by tremors, rigidity, bradykinesia, and postural instability (PD).", replace = TRUE)

glossary_add(
  term = "Idiopathic PD",
  def = "PD without identifiable genetic or environmental factors.", replace = TRUE
)

glossary_add(
  term = "Sporadic PD",
  def = "No clear family history of PD, often assumed to be idiopathic.", replace = TRUE)


glossary_add(
  term = "Alpha-synuclein",
  def = "A protein that plays a key role in the pathogenesis of Parkinson's disease, characterized by its tendency to misfold and aggregate into toxic fibrils that contribute to neurodegeneration.", replace = TRUE)

glossary_add(
  term = "SAA",
  def = "Synuclein Amplification Assay, a laboratory test used to stratify participants as either alpha-synuclein positive or negative.", replace = TRUE)

glossary_add(
  term = "Prodromal",
  def = "A period of time during which an individual experiences symptoms or warning signs of a disease before its full-blown onset.", replace = TRUE)

glossary_add(
  term = "MDS-UPDRS",
  def = "Movement Disorder Society-Unified Parkinson's Disease Rating Scale, a standardized assessment tool for evaluating motor and non-motor symptoms of Parkinson's disease.", replace = TRUE)

glossary_add(
  term = "MOCA",
  def = "Montreal Cognitive Assessment, a widely used screening test for detecting cognitive impairment.", replace = TRUE)

glossary_add(
  term = "REM",
  def = "Rapid Eye Movement sleep, a stage of sleep characterized by vivid dreams and increased brain activity and associated with PD risk when resulting in abnormal motor behaviors such as acting out dreams.", replace = TRUE)

glossary_add(
  term = "LRRK2",
  def = "Leucine-rich repeat kinase 2, a gene associated with an increased risk of developing Parkinson's disease.", replace = TRUE)

glossary_add(
  term = "GBA",
  def = "Glucocerebrosidase, a gene associated with an increased risk of developing Parkinson's disease.", replace = TRUE)

glossary_add(
  term = "PIGD",
  def = "Postural instability and gait disturbance, a subtype of Parkinson's disease characterized by postural instability and gait difficulties.", replace = TRUE)

glossary_add(
  term = "IDP",
  def = "Imaging data phenotype, a summary measurement appropriate for representing and interpreting imaging data in tabular form.", replace = TRUE)

glossary_add(
  term = "sMRI",
  def = "Structural Magnetic Resonance Imaging, a type of (usually T1-weighted) MRI that provides detailed images of brain anatomy.", replace = TRUE)

glossary_add(
  term = "dMRI",
  def = "Diffusion Magnetic Resonance Imaging, a type of MRI that measures the diffusion of water molecules in the brain, providing information on white matter tracts and brain connectivity.", replace = TRUE)

glossary_add(
  term = "rsfMRI",
  def = "Resting-state functional Magnetic Resonance Imaging, a type of functional MRI that measures brain activity while a person is at rest, providing information on brain functional connectivity.", replace = TRUE)

glossary_add(
  term = "Traveling participants",
  def = "Participants in a research study who undergo repeated assessments and evaluations at different locations or time points, often used to study accuracy and reproducibility of imaging data across scanners.", replace = TRUE)

glossary_popup("hover") 

areunique = function( x ) {
  x[ !duplicated( x[,c("studyName","commonID","age_BL","T1Hier_resnetGrade")]), ]
}


subset_and_combine_studies <- function(df, study_column, study_names, grade_column, k_values) {
  # Check that study_names and k_values have exactly 2 elements
  if (length(study_names) != 2 || length(k_values) != 2) {
    stop("study_names and k_values should each have exactly 2 elements.")
  }

  # Define a helper function to subset and select top k rows for a given study
  subset_and_select <- function(study_name, k) {
    # Subset the dataframe for the specific study
    study_df <- df[na2f(df[[study_column]] == study_name), ]
    # Select the top k rows based on the specified grade column
    selected_rows <- topk(study_df, grade_column, k)
    # Return the subset of the dataframe with the selected rows
    return(study_df[selected_rows, ])
  }

  # Apply the helper function to each study and combine the results
  study1_subset <- subset_and_select(study_names[1], k_values[1])
  study2_subset <- subset_and_select(study_names[2], k_values[2])

  # Combine the two subsets
  combined_df <- rbind(study1_subset, study2_subset)
  
  return(combined_df)
}

 
makeqcpc <- function(demog, modality='T1w') {
  qcnames <- NULL
  coreqcnames =  c("EVR", 'snr', 'psnr', 'ssim', 'cnr', "reflection_err", 'mi', 'noise', 'msk_vol', 'spc0', 'spc1', 'spc2' )
  if (modality == 'T1w') {
    qcnames = c("T1Hier_templateL1", "T1Hier_resnetGrade", 'T1Hier_mhdist', coreqcnames )
  } else if (modality == 'DTI') {
    temp = paste0( modality,"1_",coreqcnames)
    qcnames = c("T1Hier_templateL1", "T1Hier_resnetGrade", temp)
  } else if (modality == 'rsf') {
    temp = paste0( modality,"1_",coreqcnames)
    qcnames = c("T1Hier_templateL1", "T1Hier_resnetGrade", temp)
  } else {
    stop("Unsupported modality")
  }
  
  # Identify rows with any NA values in the selected columns
  na_rows <- apply(demog[, qcnames], 1, function(x) any(is.na(x)))
  
  # Filter out rows with NA values
  non_na_demog <- demog[!na_rows, qcnames]
  pcqcmat <- data.matrix(non_na_demog)
  
  # Perform PCA
  pca_result <- prcomp(pcqcmat, scale. = TRUE)
  new_pca_columns <- pca_result$x[, 1:4]
  colnames(new_pca_columns) <- paste0(modality, colnames(new_pca_columns))
  colnames(new_pca_columns)=gsub("PC","QC",colnames(new_pca_columns))
  # Create a data frame with new PCA columns filled with NA for the original rows with NA
  pca_columns_full <- matrix(NA, nrow = nrow(demog), ncol = 4)
  colnames(pca_columns_full) <- colnames(new_pca_columns)
  pca_columns_full[!na_rows, ] <- new_pca_columns
  
  # Combine the original data frame with the new PCA columns
  demog <- cbind(demog, pca_columns_full)
  
  return(demog)
}


mypartrate=0.8
npd=ngpd=nprepd=0
mvanvp3=mvanvpu=mvanvp2=mvanvp1=1
doadj=FALSE
dxcts=data.frame()
rdir=path.expand('~/code/multidisorder/data/')
rdir=path.expand('~/Downloads/ppmi_pym_data/')
ppmitrimfn = paste0(rdir,'ppmi_idps_trim_v1.4.0_',myrez,'.csv')
rgT1 = c( 5000, 2150, 950 )
########
replaceOG=TRUE
if ( file.exists( ppmitrimfn ) & ! exists("ppmitrim")  ) {
  ppmitrim = read.csv(ppmitrimfn)
  adjnames = getNamesFromDataframe( "_adjusted", ppmitrim )
  ognames = gsub("_adjusted", "", adjnames )
  if ( replaceOG ) {
    ppmitrim[,ognames]=ppmitrim[,adjnames]
  }
} else if ( ! exists("ppmi") & !exists("ppmitrim") ) {
  if ( ! exists("pymf") ) {
    v='1.4.0'
    pymfns = data.frame(
      fns=c( 
        paste0( rdir, 'ppmi_matched_qc_mm_processedCSV_v',v,'.csv'),
        paste0( rdir, 'ppmi_matched_qc_mm_processedCSVSRFIRST_v',v,'.csv'),
        paste0( rdir, 'ppmi_matched_qc_mm_processedCSVSRNext_v',v,'.csv')  ),
      rez=c("OR","SRF","SRP") )
    print(paste("ANTsPyMM Pipeline : ", myrez ) )
    pym = read.csv(  pymfns$fn[pymfns$rez == myrez ] )
    pymf = pym[ , !(colnames(pym) %in% getNamesFromDataframe("u_hier_id", pym))] # junk
    pymf = filterNAcolumns( pymf, 20 ) # remove columns with > 20% missing
    pymf[ is.na(pymf$mrimfg), 'mrimfg' ]='Unk'
    pymf[ is.na(pymf$mrimodel), 'mrimodel' ]='Unk'
    rm(pym)
    }
  if ( ! exists("adni") ) {
    adni = read.csv( path.expand( paste0(rdir, "ADNIMERGE_10Feb2024_antspymm_v",v,".csv" ) ) )
#    adnihq = highestQualityRepeat( adni, 'PTID', 'date', 'T1Hier_resnetGrade' )
  }
  studynames=c("ADNI", "NIFD4RTNI", "PPMI","UKBB","A4")
  dxnames= c("DX_bl", "DX", "DXSub", "DX", "DX" )
  sexnames=c("PTGENDER","SEX","SEX","sex_f31_0_0","PTGENDER")
  agenames=c("AGE","Age_BL","age_BL","subjectAge_BL","PTAGE")
  edunames=c("PTEDUCAT","EDUCATION","EDUCYRS","age_completed_full_time_education_f845_0_0","PTEDUCAT")
  sidnames=c("PTID","LONI_ID","PATNO","eid","sid")
  viznames=c("Years.bl","deltaTime","yearsbl","Years.bl","Years.bl")
  adni$commonSex = adni$PTGENDER
  adni$commonEdu = adni$PTEDUCAT
  adni$commonID = adni$subjectID = adni$PTID
  adni$age_BL = adni$age = adni$AGE
  adni$yearsbl = adni$Years_bl
  adni$joinedDX = adni$DX_bl
  adni$joinedDX[ grep("MCI",adni$joinedDX)]="MCI"
  adni$joinedDX[ adni$joinedDX == ""]=NA
  adni$studyName='ADNI'
  adniasynfn = path.expand( paste0(rdir,"AMPRION_ASYN_SAA_13Nov2023.csv"))
  adniasyn = read.csv( adniasynfn )
  names( adniasyn )[ names(adniasyn) == 'Result' ]='AsynStatus'
  adniasyn[ subtyper::fs(adniasyn$AsynStatus %in% c('Detected-1','Detected-2')) , 'AsynStatus'] = 'Positive'
  adniasyn[ subtyper::fs(adniasyn$AsynStatus %in% c('Not_Detected','Indeterminate')), 'AsynStatus' ] = 'Negative'
  # map 
  adni$AsynStatus=NA
  uids = unique( adniasyn$RID[!is.na(adniasyn$AsynStatus)])
  for ( k in 1:nrow(adniasyn) ) {
        losel = subtyper::fs( adni$RID == adniasyn$RID[k] )
        if ( sum(losel) > 0 ) {        
            adni$AsynStatus[ losel ] = adniasyn$AsynStatus[k]
        }
    }
  ########### PPMI critical stuff here ###########
  demog=read.csv(paste0(rdir,"PPMI_20240401/Demographics_20240401.csv"))
  curfn=paste0(rdir,'PPMI_Curated_Data_Cut_Public_20240129.csv')
  saafn = paste0(rdir,'PPMI_CSFSAA_09Oct2023/PPMI_CSFSAA_09Oct2023.csv')
  if ( ! file.exists( saafn ) ) {
    saafn = paste0(rdir,'PPMI_CSFSAA_09Oct2023.csv')
    }
  saa=read.csv( saafn )
  ppmidemog0=read.csv( curfn )
  clin2b = merge_ppmi_imaging_clinical_demographic_data(
    demog, ppmidemog0, pymf, pymversion=myrez, saa )
  # add tracker
  tracker = read.csv(paste0(rdir,"ppmi_2_0_internal_scan_tracking_report_v12-2024-02-08.csv"))

  tracker = tracker[ tracker$Modality == "MR", ]
  tracker$EVENT_ID=tracker$Visit
  tracker$EVENT_ID[ grep("Baseline",tracker$EVENT_ID)]="BL"
  tracker$date=NA
  momap=data.frame( num=c( paste0("0",1:9), "10", "11", "12") )
  rownames(momap)=c("JAN","FEB","MAR","APR","MAY","JUN","JUL","AUG","SEP","OCT","NOV","DEC")
  for ( k in 1:nrow(tracker) ) {
    fixedev=unlist(strsplit( tracker$Visit[k], "/"))
    if ( length( fixedev ) > 1 ) fixedev=fixedev[length( fixedev )]
    tracker$EVENT_ID[k]=fixedev
    fixeddt = unlist(strsplit( tracker$Scan.Date[k],"-"))
    fixeddt[2] = momap[ fixeddt[2], 'num' ]
    fixeddt = paste0( fixeddt[3] , fixeddt[2], fixeddt[1] )
    tracker$date[k]=fixeddt
    }
  trackersub = tracker[, c("Subject.ID","date" )]
  trackersub$imaging_protocol=2
  names(trackersub)=c("PATNO","date", "imaging_protocol" )
  clin2b=merge( clin2b, trackersub, by=c("PATNO","date"), all.x=TRUE )
  clin2b$imaging_protocol[ is.na(clin2b$imaging_protocol) ]=1
  # add LRRK2 
  lrrk2 = read.csv( paste0(rdir,'CSFSAA_LRRK2_MRI_SAMPLES_08Jan2024/Data-Table1.csv'))
  lrrk2$SAAGroup[ lrrk2$SAAGroup == " " ]=NA
  clin2b$LRRK2_study=NA
  uids = lrrk2$PATNO
  lrrk2cols = colnames(lrrk2)
  lrrk2cols=lrrk2cols[ !(lrrk2cols %in% "PATNO") ]
  newcols = paste0("LRRK2_", lrrk2cols )
  clin2b[,newcols]=NA
  for ( k in 1:length(uids) ) {
    clin2bsel = clin2b$PATNO == uids[k]
    if ( sum( clin2bsel ) > 0 ) {
      # print(paste("have",u))
      clin2b[clin2bsel,newcols]=lrrk2[k,lrrk2cols]
    } # else print(paste("miss",u))
  }
  clin2b$LRRK2_study = !is.na( clin2b$LRRK2_cohort)
  summary( lm( LRRK2_MRIPROTOCOL ~ DTI_dti_tsnr_dwi_mean +T1Hier_evratio + noise + snr + msk_vol,data=clin2b) )
  summary( lm( LRRK2_MRIPROTOCOL ~ DTI_dti_tsnr_b0_mean+DTI_dti_ssnr_b0_mean + DTI_dti_tsnr_dwi_mean+DTI_dti_ssnr_dwi_mean+T1Hier_resnetGrade +T1Hier_evratio + cnr + noise,data=clin2b) )
  summary( lm( LRRK2_MRIPROTOCOL ~  cnr + noise  + msk_vol + spc0+	spc1+	spc2,data=clin2b) )
  clin2b$subjectID = as.character(clin2b$subjectID)
  clin2b$imageID = as.character(clin2b$imageID)
  clin2b$dtid1 = as.character(clin2b$dtid1)
  clin2b$dtid2 = as.character(clin2b$dtid2)
  clin2b$rsfid1 = as.character(clin2b$rsfid1)
  clin2b$rsfid2 = as.character(clin2b$rsfid2)
  clin2b=dplyr::bind_rows( clin2b, adni )
  nna=!is.na( clin2b$AsynStatus )
  clin2b$DXSubAsyn[nna]=paste0( clin2b$joinedDX[nna], clin2b$AsynStatus[nna] )
  clin2b$DXSubAsyn[ clin2b$DXSubAsyn == "NANegative"]=NA
  clin2b$commonEdu[ clin2b$commonEdu == 1] = 12
  clin2b$commonEdu[ clin2b$commonEdu == 2] = 16
  clin2b$commonEdu[ clin2b$commonEdu == 3] = 20
  ppmi = clin2b
  rm(clin2b)
  rm(pymf)
  rm(adni)
  gc()
  ppmi$T1Hier_midbrain_pons_ratio = ppmi$T1Hier_vol_midbrainbrainstem / ppmi$T1Hier_vol_ponsbrainstem
  ppmi$SITE = factor( ppmi$SITE )
  ppmi$brainVolume = ppmi$T1Hier_vol_hemisphere_lefthemispheres + ppmi$T1Hier_vol_hemisphere_righthemispheres
  ppmi$brainVolume = ppmi$brainVolume / mean(ppmi$brainVolume)
  ppmi$imaging_protocol[ ppmi$studyName=='ADNI']=3
  ppmi$imaging_protocol=factor(ppmi$imaging_protocol)
  ppmi[ ppmi$studyName=='ADNI', 'age_BL']=ppmi[ ppmi$studyName=='ADNI', 'AGE']
}



```



```{r, results='asis',echo=FALSE,warning=FALSE}
glossary_style(color = "purple", 
               text_decoration = "underline",
               def_bg = "#333",
               def_color = "white")
```



```{r ppmitrim,echo=FALSE,eval=!exists("ppmitrim"),fig.width=16,fig.height=10,message=FALSE,warning=FALSE,cache=FALSE}
ppmi = fillBaselineColumn( ppmi,
        c(  'brainVolume', 'T1Hier_resnetGrade' ),
        'commonID', 'yearsbl', 0, 
        fast=T, verbose=F )[[1]]
ppmi = antspymm_predictors( ppmi, TRUE )
ppmi$maxtominspc = apply(  ppmi[,c("spc0","spc1","spc2")], FUN=max, MARGIN=1 )/
  apply(  ppmi[,c("spc0","spc1","spc2")], FUN=min, MARGIN=1 )
#############
mith=-0.5
ssimth=0.3
psnrth=12.0
mmspc=2.0
evThresh=c(0.45,0.9)  # 0.3
evThresh=c(0.0,1.0)  # 0.3
rgThresh=0.2
eps=0.0
qvary = c(eps,1.0-eps)
if ( myrez == 'SRF' ) {
  snrth=c( quantile(ppmi$snr,qvary,na.rm=T) )
} else {
  snrth=c( quantile(ppmi$snr,qvary,na.rm=T) )
}
asel = subtyper::fs( ppmi$studyName == 'ADNI' & 
  ppmi$T1Hier_resnetGrade_BL >= rgThresh & ppmi$T1Hier_resnetGrade >= rgThresh )
studies = c('PPMI','ADNI')
# studies = c('PPMI')
psel = subtyper::fs( 
    ppmi$studyName %in% studies & !(ppmi$subjectID == "101021" ) & 
    ppmi$SITE != 76 & 
    ppmi$yearsbl <= 4.5 & # limit non-random bias
    ( ppmi$snr >= snrth[1] & ppmi$snr <= snrth[2] ) &
    ppmi$maxtominspc <= mmspc & 
    ppmi$T1Hier_resnetGrade_BL >= rgThresh &
    ppmi$T1Hier_resnetGrade >= rgThresh &
#    ppmi$mi < mith & ppmi$ssim > ssimth & ppmi$psnr > psnrth &
    ppmi$EVR >= evThresh[1] &  ppmi$EVR <= evThresh[2]
)
ppmitrim = ppmi[ psel , ]
table( ppmitrim$studyName )

for ( mymo in c("T1w","DTI","rsf") ) {
  ppmitrim=makeqcpc( ppmitrim, mymo )
}

ppmitrim$mriMagneticFieldStrength[ 
  ppmitrim$mriMagneticFieldStrength %in% c(1.494)]=1.5
ppmitrim$mriMagneticFieldStrength[ 
  ppmitrim$mriMagneticFieldStrength %in% c(2.89362)]=3
ppmitrim$mriMagneticFieldStrength[ 
  !(ppmitrim$mriMagneticFieldStrength %in% c(1.5,3))]='other'
ppmitrim$mriMagneticFieldStrength=factor(ppmitrim$mriMagneticFieldStrength)

nmnames = getNamesFromDataframe( c("NM2DMT_NM","_substantianigra"), ppmitrim, exclusions='z_coord'  )
nmqcnames = getNamesFromDataframe( "NM2DMT", ppmitrim, exclusions='_substantianigra' )

```

```{r ppmitrimX,echo=FALSE,eval=TRUE,fig.width=16,fig.height=10,messages=FALSE,warning=FALSE,cache=FALSE}
library(GGally)
qcnames=c( "T1Hier_resnetGrade", "noise", "snr", "EVR", "psnr", "ssim", "mi", "msk_vol", "reflection_err", "mriPixelBandwidth", "mriSAR", "maxtominspc",'updrs_totscore' )
my_bin <- function(data, mapping, ..., low = "#132B43", high = "#56B1F7") {
  ggplot(data = data, mapping = mapping) +
    geom_bin2d(...) +
    scale_fill_gradient(low = low, high = high)
}
ssel = subtyper::fs(ppmitrim$studyName == 'PPMI')
qcdf=ppmitrim[  ssel, c(qcnames,'joinedDX')]
qcdf$snc=ppmitrim$T1Hier_vol_mtg_sn_snc_LRAVGcit168[ssel]
# qcdf=ppmi[,c(qcnames,'joinedDX')]
# qcdf$snc=ppmi$T1Hier_vol_mtg_sn_snc_LRAVGcit168
qcdf$DX = "Other"
qcdf$DX[ grep("PD",qcdf$joinedDX)]="PD"
qcdf$DX[ grep("CN",qcdf$joinedDX)]="CN"
qcdf$DX[ grep("Prod",qcdf$joinedDX)]="AR"
qcpairs <- ggpairs( qcdf, columns=c(qcnames,"snc"),
  mapping = aes(color = DX),
  lower = list(
    combo = wrap("facethist", binwidth = 1),
    continuous = wrap(my_bin,  high = "red")
  ) )
# print( qcpairs )
# gc()

# ntokeep = nrow(ppmitrim)-round(0.01*nrow(ppmitrim))
# ppmitrim = ppmitrim[ head( order(ppmitrim$T1Hier_resnetGrade,decreasing=T),ntokeep) , ]
updnames = getNamesFromDataframe("upd",ppmitrim,exclusions=c("BL","delta"))
msnames = getNamesFromDataframe("pm_",ppmitrim,exclusions=c("BL","delta"))
lrrk2names= getNamesFromDataframe("LRRK2_",ppmitrim,exclusions=c("BL","delta","LRRK2_MRIPROTOCOL"))
t1namesctx = getNamesFromDataframe( c("T1Hier","LRAVG",'dktcortex','thk'), ppmitrim)
t1namescit = getNamesFromDataframe( c("T1Hier","LRAVG",'deep','vol'), ppmitrim)
t1namesnbm = getNamesFromDataframe( c("T1Hier","LRAVG",'nbm','vol'), ppmitrim)
t1namesbst = getNamesFromDataframe( c("T1Hier",'brainstem','vol'), ppmitrim, exclusions=c("tissues","lobes"))[-1]
t1namescrb = getNamesFromDataframe( c("T1Hier",'cerebellu','LRAVG','vol'), ppmitrim, exclusions=c("tissues","lobes","dktregion"))
t1namesctxa = gsub("LRAVG","Asym",t1namesctx)
t1namescita = gsub("LRAVG","Asym",t1namescit)
t1namesnbma = gsub("LRAVG","Asym",t1namesnbm)
t1namescrba = gsub("LRAVG","Asym",t1namescrb)
t1vars = c(t1namesbst,t1namesctx,t1namescit,t1namesnbm,t1namescrb,t1namesctxa,t1namescita,t1namesnbma,t1namescrba)
ppmitrim0 = ppmitrim[ , c("subjectID","age_BL",'commonSex', "hy_BL", "duration_yrs","LEDD","MOCA",updnames, msnames, 'joinedDX','AsynStatus', "DXSubAsyn", "educ","race",'tau','ptau','abeta', 'yearsbl', 'EVENT_ID', 'date', lrrk2names,  'filename', 'brainVolume', t1vars, 'imaging_protocol', 'T1Hier_resnetGrade' ) ]

# get numbers for abstract
viztbl = table( ppmitrim0$joinedDX, ppmitrim0$EVENT_ID )
isbl=ppmitrim0$yearsbl==0
isnotbl = ppmitrim0$yearsbl >= 0.5
npd=viztbl["PDSporadic","BL"]
npdf=sum(viztbl["PDSporadic",c("V04","V06","V08","V10","V12")])
ngpd=sum( subtyper::fs(ppmitrim0$joinedDX[isbl] %in% c('PDGBA','PDLRRK2','PDSNCA' )))
nctl=viztbl["CN","BL"]
nprepd=sum( subtyper::fs(ppmitrim0$joinedDX[isbl] %in% c('ProdromalGBA','ProdromalLRRK2' ,'ProdromalSNCA','ProdromalSporadic' )))
vizfll=mean( ppmitrim0$yearsbl[isnotbl] )
vizfll=insight::format_value( vizfll )
vizfll2=IQR( ppmitrim0$yearsbl[isnotbl] )
vizfll2=insight::format_value( vizfll2 )
quartiles <- quantile(ppmitrim0$yearsbl[isnotbl], probs = c(0.25, 0.75))
quartilesf=insight::format_value(quartiles)
# tt = table( table( ppmitrim0$subjectID[!isbl] ) )

if ( FALSE ) {

ppmitrimbl=ppmitrim[ ppmitrim$EVENT_ID=='BL', ]
vv='T1Hier_vol_bn_str_pu_Asymdeep_cit168'
vv='T1Hier_vol_bn_gp_gpi_Asymdeep_cit168'
vv='T1Hier_vol_bn_gp_gpi_LRAVGdeep_cit168'
qq=quantile( ppmitrimbl[,vv], 0.95 )
table( ppmitrimbl$DXSubAsyn[ ppmitrimbl[,vv] > qq ] )
qq=quantile( ppmitrimbl[,vv], 0.025 )
table( ppmitrimbl$DXSubAsyn[ ppmitrimbl[,vv] < qq ] )

# convert Asym to AsymNorm 
ppmitrim = antspymm_predictors( ppmitrim, TRUE )
asymvars = getNamesFromDataframe( "Asym", ppmitrim )
for ( x in asymvars ) {
  y=gsub("Asym","LRAVG",x)
  myasym = ppmitrim[,x]
  ppmitrim[,x]=myasym/(ppmitrim[,y]*2)
  if ( max(ppmitrim[,x],na.rm=T) > 1 ) derka
}
}
doadj=!("DTI_mean_md_nbm_LRAVG_pos_adjusted" %in% colnames(ppmitrim))
########
```


```{r ppmitrimAdj,echo=FALSE,eval=doadj,fig.width=16,fig.height=10,message=FALSE,warning=FALSE,cache=FALSE}
ppmitrim = ppmitrim[ ppmitrim$mrimfg %in% c("GE","Philips","Siemens"), ]
nmstotest2=antspymm_predictors( ppmitrim, TRUE, TRUE )
nmstotest2=nmstotest2[ -multigrep( c( "T1w_mean", "alff","peraf","template","fcnxpro134","fcnxpro122","_adjusted"), nmstotest2)]
# adjust each of these vars
ppmitrim$isTrain=FALSE
# fairly balanced
dxsel = ppmitrim$joinedDX %in% c("CN",'ProdromalLRRK2','ProdromalGBA','ProdromalSporadic','SWEDDSporadic')
# dxsel = ppmitrim$joinedDX %in% c("CN")
agsel = ppmitrim$age_BL >= 50 & ppmitrim$age_BL <= 75
ppmitrim$isTrain[ dxsel & agsel ]=TRUE
table(ppmitrim$isTrain, ppmitrim$studyName )
############################################
ct=0
for ( x in nmstotest2 ) {
    ct=ct+1
    if ( ct %% 100 == 0 ) cat( paste( ct / length(nmstotest2)*100 , "%" )  )
    if ( antspymm_vartype(x) %in% c('rsfMRI','DTI') ) {
        myform = paste( x, " ~ studyName + mrimfg")
    } else myform = paste( x, " ~ studyName + mrimfg + mriMagneticFieldStrength ")
    ppmitrim=adjustByCovariates( ppmitrim, myform, 'isTrain', TRUE, TRUE )
}
nmstotest2adj=paste0(nmstotest2,'_adjusted' )
doadj=FALSE
```

# Overview of the supplement

We first provide additional information on the currently processed ADNI and PPMI cohort including the genetic and prodromal subgroups along with asyn SAA status for each.  Second, we share didactic material explaining in greater detail how to interpret the plots in Figure 9 of the main text.

## The full PPMI cohort 

The main text focused on the largest groups available in this data release: controls and sporadic PD.  However, other PD groups are of growing interest for more focused research.  In particular, both GBA and LRRK2 PD appear to take a different disease trajectory than is typical of sporadic PD \footnote{Typical should probably not be used to describe PD progression due to the immense variability that is observed across individual patients.}  Moreover, the prodromal stage is of growing interest due to the combination of an increased ability to identify these subjects and the potential for early intervention.  We believe the validity arguments made in the primary text extends to the IPDs that are associated with these other groups, summarized in Table S1.


```{r tblprep,eval=TRUE,echo=FALSE,cache=FALSE,warning=FALSE}

nmstotest2=antspymm_predictors( ppmitrim, TRUE, TRUE )
nmstotest2=nmstotest2[ -multigrep( c("alff","peraf","template","fcnxpro134","fcnxpro122","_adjusted","_BL","_delta","T1w_mean","area"), nmstotest2)]

if ( ! exists("ppmitrimsave") & FALSE ) {
#  ppmitrimsave = ppmitrim
#  nmstotest2=gsub("_adjusted","",nmstotest2adj)
#  ppmitrim[,nmstotest2]=ppmitrim[,nmstotest2adj]
  write.csv( ppmitrim, ppmitrimfn, row.names=FALSE )

  simfile="~/Downloads/ppmi_pym_data/PPMISiMLRW100"
  presim=read_simlr_data_frames( path.expand(simfile), 
      c("t1","t1a", "dt", "dta", "rsf" ))
  ppmitrim = apply_simlr_matrices(  ppmitrim, presim, absolute_value=rep(TRUE,length(presim)), robust=FALSE )[[1]]
  simnames = apply_simlr_matrices(  ppmitrim, presim, absolute_value=rep(TRUE,length(presim)), robust=FALSE )[[2]]
#########
  cnames = getNamesFromDataframe("common",ppmitrim)
  cnames[ which(cnames=='commonAge')]='age_BL'
  trimnames=c("studyName",cnames,'yearsbl', "date","filename",
    getNamesFromDataframe("QC",ppmitrim), antspymm_qc_names(),nmstotest2, simnames)
  trimnames=unique(intersect(trimnames,colnames(ppmitrim)))
  write.csv( ppmitrim[,trimnames], gsub(".csv","_trim.csv",ppmitrimfn,fixed=TRUE), row.names=FALSE )
}

###
toadd = c(  "age_BL", "commonSex", 'race', 'hy_BL', 'duration_yrs', "updrs_totscore", 'AsynStatus', 'LEDD', 'subgroup', 'imaging_protocol')
toadd = c(  "age_BL", "commonSex", 'race',  'duration_yrs', "updrs_totscore", 'AsynStatus', 'LEDD', 'subgroup', 'imaging_protocol')
biomarkers = c('abeta', 'tau',  'nfl_serum', 'mean_striatum' )
tblcols = unique( c( toadd, 'joinedDX'  ) )
bsel = ppmitrim$yearsbl==0 # & ppmi$studyName == 'PPMI'
joinedem=ppmitrim[bsel,tblcols]
joinedem$imaging_protocol=as.character(joinedem$imaging_protocol)
#joinedem$hasDTI=!is.na(ppmitrim[bsel,'DTI_dti_FD_max'])
#joinedem$hasfMRI=!is.na(ppmitrim[bsel,'rsfMRI_fcnxpro129_FD_max'])
names(joinedem)[1]=tblcols[1]='age'
names(joinedem)[2]=tblcols[2]='Sex'
wsaa=which(names(joinedem)=='AsynStatus' )
names(joinedem)[wsaa]=tblcols[wsaa]='CSFSAA'
tblcols=colnames(joinedem)
tblcols=gsub("_",".",tblcols)
tblcols=gsub(".mean","",tblcols)
tblcols=gsub("DTI.","",tblcols)
tblcols=gsub("rsfMRI.fcnxpro122.","rsfMRI.",tblcols)
names(joinedem)=tblcols
joinedem$race = as.character( joinedem$race )
joinedem[ subtyper::fs(joinedem$race == '1'), 'race' ] = 'White'
joinedem[ subtyper::fs(joinedem$race == '2'), 'race' ] = 'Black'
joinedem[ subtyper::fs(joinedem$race == '3'), 'race' ] = 'Asian'
joinedem[ subtyper::fs(joinedem$race == '4'), 'race' ] = 'Other'
joinedem[ is.na(joinedem$race), 'race' ] = 'not.spec.'
joinedem$joinedDX = gsub("Prodromal",'AR',joinedem$joinedDX)
joinedem$joinedDX=as.character(joinedem$joinedDX)
joinedem$joinedDX[ 
  multigrep( c("ARSporadic"), joinedem$joinedDX )]='Sporadic.Pro'
joinedem$joinedDX[ 
  multigrep( c("ARGBA", "ARLRRK2",  "ARSNCA"), joinedem$joinedDX )]='Gen.Pro'
joinedem$joinedDX[
  multigrep( c( "PDPRKN", "PDSNCA"), joinedem$joinedDX )]='OtherGen.PD'
joinedem$joinedDX[ 
  multigrep( c("PDGBA"), joinedem$joinedDX )]='GBA.PD'
joinedem$joinedDX[
  multigrep( c("PDLRRK2"), joinedem$joinedDX )]='LRRK2.PD'
joinedem$joinedDX[ 
  multigrep( c("PDSporadic"), joinedem$joinedDX )]='Sporadic.PD'
mlevs = c( "CN", "Gen.Pro", "Sporadic.Pro","OtherGen.PD", 'LRRK2.PD', 'GBA.PD', "Sporadic.PD")
joinedem$joinedDX = factor( joinedem$joinedDX, levels = mlevs )
mytbl = joinedem[ !is.na(joinedem$joinedDX),c(tblcols)]
mytbl$imaging.protocol=as.character(mytbl$imaging.protocol)
mytbl$imaging.protocol[ mytbl$imaging.protocol=='1']='PPMI1'
mytbl$imaging.protocol[ mytbl$imaging.protocol=='2']='PPMI2'
mytbl$imaging.protocol[ mytbl$imaging.protocol=='3']='ADNI'
rownames(mytbl)=1:nrow(mytbl)
mycap="Table S1. Baseline PPMI IDP cohort. Pro=prodromal; MSA = multi-system atrophy; OtherGen = other genetic; RBD = REM Sleep Behavior Disorder."
####
```


```{r clusterCharARtbl,eval=TRUE,echo=FALSE,cache=FALSE,results='asis',warning=FALSE}
sttblPD=ztable(mytable(joinedDX~.,data=mytbl),zebra=2)

# Print the ztable object, ensuring it's treated as LaTeX code
print(sttblPD, type = "latex",size=3,caption=mycap,sidewaystable=TRUE)


```


\LaTeX{
\setcounter{table}{1}
\renewcommand{\thetable}{S\arabic{table}}
}


\newpage


```{r pkgs,echo=FALSE,eval=TRUE,message=FALSE,cache=FALSE,warning=FALSE}
set.seed( 0 )
library(plyr)
library(ggplot2)
library( GoodmanKruskal )
library(forcats)
library(heatmap3)
library(permute)
library(interactions)
library(reghelper)
library(MGMM) # GMM methods 
library(WeightedCluster) # cluster quality 
library(tidyverse)
library(clValid)
library( cluster.datasets )
library(magrittr)
library(corrplot)
library(GGally)
library(ANTsR)
library(subtyper)
library(plyr)
library(data.table)
library(ggfortify)
library(ztable)
library(moonBook)
library(kableExtra)
library(flextable)
library(plotly)
library( ztable )
library(ggpubr)
library(gridExtra)
library(ggsci)
options(digits=3)
library(lmerTest)
library(lme4)
library(mlr3verse)
library(Evacluster)
library( caret )
library( optmatch )
library( globaltest )
library(emmeans)
library(analyzer)
library(forcats)
options( ztable.type="latex" )
tblcmd = ztable
```

```{r functions,echo=FALSE,eval=TRUE,warning=FALSE,message=FALSE,cache=FALSE}

normalizersfmri = function( dd, applyit='none' ) {
  nums=c(122,129,134)
  t=c('_2_', "_falff", "_alff", "_peraf")
  tnm=c("fcnxmean" ,'falffmean', 'alffmean', 'perafmean')
  tns=c("fcnxsd" ,'falffsd', 'alffsd', 'perafsd')
  for ( k in 1:length(nums) ) {
    ct=0
    for ( tt in t ) {
      ct=ct+1
      protype=paste0("fcnxpro",nums[k])
      srch=c(t[ct],protype)
#      print( srch )
      mycolnm=getNamesFromDataframe( srch , dd )
#      print(head(mycolnm,2))
#      print(mycolnm  )
      mnnm=paste0( tnm[ct],"_",nums[k])
      sdnm=paste0( tns[ct],"_",nums[k])
#      print(paste(mnnm))
      dd[,mnnm]=apply( dd[, mycolnm], FUN=mean,MARGIN=1)
#      print(paste(sdnm))
      dd[,sdnm]=apply( dd[, mycolnm], FUN=sd,MARGIN=1)
#      print( paste( " apply ", applyit, mnnm ) )
      if ( applyit == 'mean' ) dd[,mycolnm]=dd[,mycolnm]-dd[,mnnm]
      if ( applyit == 'dividemean' ) dd[,mycolnm]=dd[,mycolnm]/dd[,mnnm]
      if ( applyit == 'meansd' ) {
        dd[,mycolnm]=(dd[,mycolnm]-dd[,mnnm])/dd[,sdnm]
      }
    }
  }
  return( dd )
}
```


```{r quicknnh,echo=FALSE,eval=TRUE,fig.width=10,fig.height=4.5,warning=FALSE,message=FALSE,fig.cap='SiMLR mapping between M3RI and PD symptomology.'}
########
#########################################################
maxy=2.25
n.comp=100
simfile=paste0("~/Downloads/ppmi_pym_data/pub_",n.comp,'_rezOR_only_ADNI')
coggersAll = c("ess", "gds", "lns", "MOCA", "NP1APAT", "NP1ANXS", "NP1DPRS", "pigd", "rem", "quip", "scopa", "stai", "VLTANIM", "updrs1_score", "updrs2_score", "updrs3_score", "updrs_totscore", 'ADAS13', 'hvlt_immediaterecall', 'FAQ', 'EcogPtTotal', 'EcogSPTotal' , 'MMSE' )
coggers = c("ess",  "MOCA", "NP1APAT", "NP1DPRS", "pigd", "rem",  "scopa",  "updrs1_score", "updrs2_score", "updrs3_score", "updrs_totscore", 'ADAS13', 'hvlt_immediaterecall', 'FAQ', 'EcogPtTotal', 'EcogSPTotal' , 'MMSE' )
coggers = c(  'MOCA',  "NP1APAT", "NP1DPRS", "pigd", "rem", "updrs1_score", "updrs2_score", "updrs3_score", "updrs_totscore" )
coggers = c(  'MOCA', "pigd", "rem", "updrs1_score", "updrs2_score", "updrs3_score", "updrs_totscore" )
presim=read_simlr_data_frames( path.expand(simfile), 
    c("t1","t1a", "dt", "dta", "rsf" ), verbose=TRUE )
npc = ncol(presim[[1]])
# npc = 10
# coggers = c( 'MOCA',  "updrs1_score", "updrs2_score", "updrs3_score"  )
mydogp = apply_simlr_matrices(  ppmitrim, presim, robust=FALSE, 
  center = TRUE, scale = TRUE,
  absolute_value=rep(TRUE,length(presim)) )[[1]]
mydogp$SITE2=factor(paste0(mydogp$SITE,mydogp$studyName))
nnhnames = apply_simlr_matrices(  mydogp, presim, robust=FALSE, absolute_value=rep(TRUE,length(presim)) )[[2]]
mydogp$roundyears = round( mydogp$yearsbl )
# mydogp$LEDD[ is.na(mydogp$LEDD) ]=0
# mydogp$duration_yrs[ is.na(mydogp$duration_yrs) ]=0
mydogp$hy[ is.na(mydogp$hy) ]=0
mydogp = fillBaselineColumn( mydogp,
        c(nnhnames,coggersAll,'hy'),
        'subjectID', 'yearsbl', 0, 
        fast=T, verbose=F )[[1]]
####################
stadnnh=data.frame()
cogglist=list()
cogglistnames=list()
nnhqth = 0.05 / ( npc * length(coggers) )
nnhqth = 1e-4
#########################
# for ( voi in c(37,52,79)  ) {
for ( voi in c(52)  ) {
  for ( cog in ('updrs_totscore') ) {
    n = nrow(stadnnh)+1
    vars3=paste0( names(presim), paste0( "PC", voi ))
    vars3f=paste("(",paste(vars3,collapse='+'),")")
    dxsel = c( "CN", "PDSporadic" )
    covarsX=" commonEdu + age_BL + commonSex + ( yearsbl ) "
    if ( cog %in% c("updrs_totscore", "updrs3_score"  , "pigd" ))
      covarsX=paste("LEDD + ",covarsX )
    redspns=paste0(cog,"_delta")
    cogformcommon=paste( redspns, "~ ",paste0(cog,"_BL"),"+(1|commonID)+(1|SITE2)+ ", covarsX )
    cogformcommon=paste( redspns, "~ ",paste0(cog,"_BL"),"+(1|commonID)+ ", covarsX )
    newform = paste( cogformcommon, " * ", vars3f )
    severity_selection = na2f( mydogp$updrs_totscore_BL <= 30 )
    severity_selection = na2f( mydogp$hy_BL %in% c(0, 1, 2) )
    dxvar='DXSubAsyn'
    dxvar='joinedDX'
    mydogpnna=mydogp[ complete.cases( mydogp[,all.vars(as.formula(newform))] ) & 
      na2f( mydogp$DTI_dti_FD_mean <= 5 & mydogp$rsfMRI_fcnxpro129_FD_mean <= 0.5 ) &
      mydogp$imaging_protocol %in% c(1,2) & # no ADNI
      (mydogp[,dxvar] %in% dxsel ) & # severity_selection &
      mydogp$yearsbl <= maxy, ]
    mydogpnna=truncatehi(mydogpnna,paste0('t1aPC',voi),t=4,removeit=FALSE)
    mydogpnna=truncatehi(mydogpnna,paste0('dtaPC',voi),t=4,removeit=FALSE)
    topkval = round( 0.985 * nrow(mydogpnna) )
    mydogpnna = mydogpnna[ topk( mydogpnna, "T1Hier_resnetGrade", topkval), ] # quality selection
    mydogpnnaBase = mydogpnna[ mydogpnna$yearsbl == 0.0, ]
    mydogpnnaLong = select_longitudinal_subjects( mydogpnna, 'commonID', 'roundyears', 2, baseline=0.0 )
    if ( n == 1 ) {
      mycogn=count_unique_subjects_per_diagnosis( mydogpnnaBase, dxvar,
      'commonID', format_counts=F )
      mycogn = paste( with(mycogn, paste(Diagnosis, Unique_Subjects, sep = ": ")), collapse=' / ' )
      mycognL=count_unique_subjects_per_diagnosis( mydogpnnaLong, dxvar,
      'commonID', format_counts=F )
      mycognL = paste( with(mycognL, paste(Diagnosis, Unique_Subjects, sep = ": ")), collapse=' / ' )
    }
    if ( cog %in% c("NP1DPRS","NP1APAT") )
      mydogpnnaLong[,redspns]=as.numeric(mydogpnnaLong[,redspns]>0)
    resultsX = lm_anv_p_and_d( mydogpnnaBase, cog, vars3f, covarsX, predictoroperator='+', verbose=F )
    anvpx=resultsX$model_comparison$Pr[2]
    if ( nrow(mydogpnnaLong) > 10 ) {
      if ( sd( mydogpnnaLong$LEDD) == 0 ) {
        newform = gsub("+ LEDD","",newform)
        newform = gsub("+ duration_yrs","",newform)
      }
      mydogpnnaLong = scale_variables_in_equation( mydogpnnaLong, newform , 
        variables_to_exclude=c('yearsbl',vars3, cog, redspns ))
      mdl=suppressMessages(lmer( newform, data=mydogpnnaLong ))
      bmdl=suppressMessages(lmer( cogformcommon, data=mydogpnnaLong ))
      nlong = length(unique( mydogpnnaLong$commonID ))
      mycoffs = coefficients( summary( mdl ) )
      anvp = suppressMessages( anova(bmdl,mdl)$Pr[2] )
      othercols=rownames(mycoffs)[ grep("PC",rownames(mycoffs))]
      chnd = data.frame( 
        effectsize::t_to_d( mycoffs[ othercols, 't value'], c(nlong,nlong) ))
      rownames(chnd)=othercols
      stadnnh[n,'cog']=cog
      stadnnh[n,'voi']=as.character(voi)
      stadnnh[n,'n.x']=resultsX$n
      stadnnh[n,'n.long']=nlong
      stadnnh[n,'anv.x']=anvpx
      stadnnh[n,'anv']=anvp
      othercolsnm=c("d.b","d.y")
      stadnnh[n,othercolsnm]=c(
        max(abs(resultsX$effect_sizes[,1])),
        max(abs(chnd[6:10,1]))
      )
      stadnnh[n,'res']=myrez
      stadnnh[n,'singular']=isSingular(mdl)
      if ( anvpx < anvp ) {
        mycoffs=( resultsX$coefficients[-1,c(3:4)]) 
        vars3xt=intersect(vars3,rownames(mycoffs))
        theorder = order(mycoffs[vars3xt,"Pr(>|t|)" ])
        mycoffsord=mycoffs[vars3xt,][theorder,]
        xviz = rownames(mycoffsord)[1]
        xviz2 = rownames(mycoffsord)[2]
        wmat = unlist( strsplit( xviz, 'PC') )
        wmat2 = unlist( strsplit( xviz2, 'PC') )
        n2show=2
        stadnnh[n,paste0("S_m1_",1:n2show)]=names(interpret_simlr_vector2( presim[[wmat[1]]] , paste0('PC',wmat[2]), n2show=n2show ))
        if ( mycoffsord[2,'Pr(>|t|)'] <= 0.01 ) {
          stadnnh[n,paste0("S_m2_",1:n2show)]=names(interpret_simlr_vector2( presim[[wmat2[1]]] , paste0('PC',wmat2[2]), n2show=n2show ))
        }
      } else {
        mycoffs=( mycoffs[-1,c(4:5)])
        vars3xt=paste0("yearsbl:",vars3)
        vars3xt=intersect(vars3xt,rownames(mycoffs))
        theorder = order(mycoffs[vars3xt,"Pr(>|t|)" ])
        mycoffsord=mycoffs[vars3xt,][theorder,]
        xviz = gsub("yearsbl:","",rownames(mycoffsord)[1])
        xviz2 =  gsub("yearsbl:","",rownames(mycoffsord)[2])
        wmat = unlist( strsplit( xviz, 'PC') )
        wmat2 = unlist( strsplit( xviz2, 'PC') )
        n2show=2
        stadnnh[n,paste0("S_m1_",1:n2show)]=names(interpret_simlr_vector2( presim[[wmat[1]]] , paste0('PC',wmat[2]), n2show=n2show ))
        if ( mycoffsord[2,'Pr(>|t|)'] <= 0.05 ) {
          stadnnh[n,paste0("S_m2_",1:n2show)]=names(interpret_simlr_vector2( presim[[wmat2[1]]] , paste0('PC',wmat2[2]), n2show=n2show ))
        }
      }
      c2viz = c( "updrs1_score48", "updrs3_score64","updrs2_score76")
      c2viz = c("NP1DPRS48", "updrs3_score55","updrs2_score50")
      c2viz = c("updrs3_score4", "MOCA33", "updrs_totscore50", "updrs_totscore55", "rem70", "NP1DPRS71") # all
      c2viz = c("updrs3_score1", "MOCA60", "updrs_totscore64", "pigd50" ,"updrs2_score97") # sporadic+cn
      c2viz = c("updrs_totscore52", "MOCA79", "updrs2_score37")
      if ( paste0(cog,voi) %in% c2viz   ) {
#      if ( paste0(cog,voi) %in% c2viz  ) {
        cogdel = paste0(cog,"_delta")
        anvp = insight::format_p( anvp, digits=4 )
        anvpx = insight::format_p( anvpx, digits=4 )
        voix=paste0("SiMLR",voi)
        ttl=( paste(  voix, myrez, " + ", cog,  " X: o", anvpx,  " L: o", anvp ) )
        m1=compare_model_predictions( resultsX$base_model, resultsX$full_model, mydogpnnaBase, cog, annot_size=3 )
        m2=compare_model_predictions( bmdl, mdl, mydogpnnaLong, redspns, annot_size=3 )
#        grid.arrange( grobs=m1, nrow=2 )
#        grid.arrange( grobs=m2, nrow=2 )
#        for ( k in 1:length(vars3)) 
#          print( jtools::effect_plot( resultsX$full_model,  !!sym(vars3[k]),  
#                interval=T, facet.modx=T, plot.points = F, partial.residuals=FALSE ) + ggtitle(paste("X:",vars3[k])) + #theme(legend.position = "none")  )
    }
  }
}
}
#####################
```

\newpage


## Understanding interaction effect plots

Interaction plots, such as those generated by the **`interactions`** package, help visualize the relationship between two or more predictor variables in a regression model, highlighting how the effect of one variable (the "moderator") changes depending on the value of another (the "focal" variable). In the manuscript, we employ the `interact_plot()` function to plot the interaction between **`yearsbl`** and a given IDP. These plots include confidence intervals to show the precision of the estimates.

The first plot faceted by the moderator (**`facet.modx = TRUE`**) presents multiple small plots, each corresponding to a different level of the moderator variable, allowing the reader to examine the interaction within each subgroup separately. This is helpful when the effect of the predictor might differ substantially across levels of the moderator. The second plot, with **`facet.modx = FALSE`**, integrates the interaction in a single plot, showing how the interaction unfolds across the entire range of the moderator. The absence of plotted points and partial residuals ensures that the reader focuses solely on the fitted lines and their confidence intervals, making it easier to interpret the overall trend without distraction from individual data points or noise.

The reader should interpret these interaction plots by comparing how the slope of **`yearsbl`** (the focal predictor) changes depending on different values of the moderator variable. A steeper or flatter slope in different facets (or sections of the integrated plot) suggests that the relationship between **`yearsbl`** and the outcome variable varies depending on the moderator, implying a significant interaction effect.  The example shown below focuses on just one of the SiMLR IDPs and how it relates to the MDS-UPDRS total score.  As in the main text, the "omnibus" effect is gained by using `anova` to comare a baseline model (no IDPs) to an extended model with IDPs.   The $p$-value for this example is `r insight::format_p(anova(bmdl,mdl)$Pr[2])` which indicates that the additional IDP variables improve the model fit significantly.  The coefficients for the base model are shown in Table S2.  The coefficients for the extended model are shown in Table S3.  Investigating the invididual $p$-values in the extendend model (Table S3) suggests that the asymmetry in the DTI-derived variables (dtaPC52) and the resting state connectivity variables (rsfPC52) are most responsible for this improvement.   Figure 1 illustrates this visually for dtaPC52 while Figure 2 shows rsfPC52.  The remainder of the effect plots in the main manuscript can be interpreted in the same or a similar manner.


```{r star,echo=FALSE,results='asis'}
library( stargazer )
library( broom.mixed )
btidy=broom.mixed::tidy(bmdl, effects = "fixed")
itidy=broom.mixed::tidy(mdl, effects = "fixed")
btidy=btidy[,c("term","estimate","statistic","df","p.value")]
itidy=itidy[,c("term","estimate","statistic","df","p.value")]

kable(btidy, format = "latex", booktabs = TRUE, digits = 3, 
      caption = "Fixed effects from the baseline mixed effects model of total MDS-UPDRS score") %>%
  kable_styling(latex_options = "striped", stripe_color = "gray!22")

kable(itidy, format = "latex", booktabs = TRUE, digits = 3, 
      caption = "Fixed effects from the extended (baseline + SiMLR IDP) model of total MDS-UPDRS score") %>%
  kable_styling(latex_options = "striped", stripe_color = "gray!22")

# Display fixed effects using stargazer
# stargazer(as.data.frame(btidy), type = "latex", summary = FALSE,
#          title = "Fixed Effects from base Model", digits = 3)

# Display fixed effects using stargazer
# stargazer(as.data.frame(itidy), type = "latex", summary = FALSE,
#          title = "Fixed Effects from extended (base + IDP) Model", digits = 3)

```



```{r theplotsDTA,echo=FALSE,eval=TRUE,fig.width=8,fig.height=4.5,warning=FALSE,message=FALSE,fig.cap='SiMLR mapping between M3RI and PD symptomology: DTA. The effect plot for the interaction of dtaPC52 with time shows its influence of UPDRS total score.  The top plot shows the faceted version of the plot on the bottom.  I.e. the same information is in both plots.   We see that increased asymmetry in these dMRI-derived variables leads to increased change in the score over time.'}
k=4
gg0=interactions::interact_plot( mdl,  'yearsbl', !!sym(vars3[k]), interval=T, facet.modx=T , plot.points = F, partial.residuals=FALSE,linearity.check = FALSE ) + ggtitle(paste("L:",vars3[k], ' faceted '))  + theme(legend.position = "none")
gg1=interactions::interact_plot( mdl,  'yearsbl', !!sym(vars3[k]), interval=T, facet.modx=FALSE , plot.points = F, partial.residuals=FALSE,linearity.check = FALSE ) + ggtitle(paste("L:",vars3[k], ' integrated '))  + theme(legend.position = "none")

grid.arrange( grobs = list(gg0, gg1 ) )

```




```{r theplotsRSF,echo=FALSE,eval=TRUE,fig.width=8,fig.height=4.5,warning=FALSE,message=FALSE,fig.cap='SiMLR mapping between M3RI and PD symptomology: rsfMRI. The effect plot for the interaction of dtaPC52 with time shows its influence of UPDRS total score.  The top plot shows the faceted version of the plot on the bottom.  I.e. the same information is in both plots.   We see that increased connectivity in these rsfMRI-derived variables leads to increased change in the score over time.'}
k=5
gg0=interactions::interact_plot( mdl,  'yearsbl', !!sym(vars3[k]), interval=T, facet.modx=T , plot.points = F, partial.residuals=FALSE,linearity.check = FALSE ) + ggtitle(paste("L:",vars3[k], ' faceted '))  + theme(legend.position = "none")
gg1=interactions::interact_plot( mdl,  'yearsbl', !!sym(vars3[k]), interval=T, facet.modx=FALSE , plot.points = F, partial.residuals=FALSE,linearity.check = FALSE ) + ggtitle(paste("L:",vars3[k], ' integrated '))  + theme(legend.position = "none")

grid.arrange( grobs = list(gg0, gg1 ) )

```




## The full SiMLR feature sets for these predictors 


We can also show -- here -- with this additional space the full feature sets associated with these SiMLR IDPs.  Figure 3 shows the asymmetry related features.  Figure 4 shows the resting state features.  A higher weight indicates a larger contribution to the final predictor value.  Note that weights are unsigned and therefore represent an average of these variables.  The maximum weight for each feature vector is 1 and we display only those features with individual weights greater than or equal to a given threshold.  The explicit `ANTsR` code for these displays is shown.


```{r simlridpDTA,echo=FALSE,fig.width=10,fig.height=7,echo=TRUE}
wmat=c("dta",52)
v0=interpret_simlr_vector2( presim[[wmat[1]]] , paste0('PC',wmat[2]), n2show=50 )
v0=v0[v0>0.01]
todisp=data.frame( feature=names(v0), weight=v0 )
ggbarplot(todisp,'feature','weight') + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 80, hjust = 1))

```


```{r simlridpRSF,fig.width=18,fig.height=10,echo=FALSE}
wmat=c("rsf",52)
v1=interpret_simlr_vector2( presim[[wmat[1]]] , paste0('PC',wmat[2]), n2show=80 )
v1=v1[v1>0.1]
todisp=data.frame( feature=names(v1), weight=v1 )
ggbarplot(todisp,'feature','weight') + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 80, hjust = 1))

```